---
title: "Tornado"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: fill
runtime: shiny
---

```{r setup, include=FALSE}
library(flexdashboard)
library(ggplot2)
library(dplyr)
library(tidytuesdayR)
library(tidyverse)
library(here)
library(ggpubr)
library(lubridate)
library(forcats)
library(purrr)
```

```{r global}
if (!dir.exists(here("data"))) {
    dir.create(here("data"))
}

files = c("tornados.rda")
if (any(!file.exists(here("data", files)))) {
    tuesdata = tidytuesdayR::tt_load('2023-05-16')
    tornados = tuesdata$tornados
    # save the files to rda objects
    save(tornados, file = here("data", "tornados.rda"), compress = "xz")
}

load(here("data", "tornados.rda"))
```


# About 

## Column {data-width=650 }

### Video illustration        

<iframe
    width="640"
    height="480"
    src="https://youtu.be/XmbQ8PgmX_4"
    frameborder="0"
    allow="autoplay; encrypted-media"
    allowfullscreen
>
</iframe>

### Purpose of this dashboard 

This dashboard aims to provide information of tornadoes in the United State from 2012-2022. Specifically, include the 
maginitude, injuries statistics of each state, as well as the trend of the Tornado.


### Data Source  

The data is from [Tidytuesday](https://github.com/rfordatascience/tidytuesday/blob/master/data/2023/2023-05-16/readme.md). And originally from comes from NOAA's National Weather Service Storm Prediction Center Severe Weather Maps, Graphics, and Data Page. 

# Additional Info 

Column {data-width=350,  .tabset .tab-fade}
-----------------------------------------------------------------------

### More about Dataset 

Variable in the original dataset but not include in the analysis       
*   om: Tornado number. Effectively an ID for this tornado in this year.       
*   dy: Day of the month         
*   time: Time           
*   tz: Canonical tz database timezone.        
*   datetime_utc: Date and time normalized to UTC        
*   stf: State FIPS (Federal Information Processing Standards) number          
*   fat: Number of fatalities         
*   loss: Estimated property loss information in dollars.         
*   slat: Starting latitude in decimal degrees.         
*   slon: Starting longitude in decimal degrees 
*   elat: Ending latitude in decimal degrees.         
*   elon: Ending longitude in decimal degrees          
*   len: Length in miles          
*   wid: Width in yards           
*   ns: Number of states affected by this tornado           
*   sn: State number for this row. 1 means the row contains the entire track information for this state, 0 means there is at least one more entry for this state for this tornado           
*   f1: FIPS code for the 1st county           
*   f2: FIPS code for the 2nd county.    
*   f3: FIPS code for the 3rd county.           
*   f4: FIPS code for the 4th county.           
*   fc: Was the mag column estimated?

### More about Tornado    

More Information about Tornado

About Tornado       
[Here](https://www.weather.gov/ffc/torntext) is a intoduction of the Tornado, if you are interested in. 

Magnitude        
Magnitude on the F scale (EF beginning in 2007). More info can be find [here](https://www.weather.gov/oun/efscale)

[State FIPS (Federal Information Processing Standards) number](https://www.bls.gov/respondents/mwr/electronic-data-interchange/appendix-d-usps-state-abbreviations-and-fips-codes.htm)

Data 
=====================================

Column {data-height="150" data-width="75" }
-------------------------------------


### Data preview 
The data is from Tidytuesday. And originally from comes from NOAA's National Weather Service Storm Prediction Center Severe Weather Maps, Graphics, and Data Page. 

```{r}
tornados_shiny=tornados%>%
    filter(yr>=2012)%>%
    filter(mag>0)%>%
    arrange(date)%>%
    select(yr,mo, date, st, mag, inj)%>%
    rename(year=yr, month=mo, state=st, magnitude=mag, injuries=inj)%>%
    na.omit()
DT::datatable(tornados_shiny)
```


# Monthly Data

## Column { .sidebar }

### Year(2012-2020)

```{r}
selectInput(inputId = "state_selected", 
              label = "Select a state to see the distribution of tornado in magnitude and injuries:",
              choices = tornados_shiny%>% 
              pull(state) %>% 
              unique()%>%
              sort(), 
              selected = "AL")
selectInput(inputId = "year_selected", 
              label = "Select a year to see the distribution of tornado in magnitude and injuries:",
              choices = tornados_shiny%>% 
              pull(year) %>% 
              unique()%>%
              sort(), 
              selected = "2012")
```

## Row 
### Tornados magnitude {data-width=550}

```{r, fig.width= 7, fig.height=6}
renderPlot({
state_yr = tornados_shiny %>% 
  filter(state == input$state_selected & year==input$year_selected)

state_yr%>%
    ggplot(aes(x=month, y=magnitude))+
    geom_point()+
    labs(title = "Tornados magnitude",
     x="Month", y="tornados magnitude", caption = "Chuyi Sun")
})
```

### Tornados injuries 

```{r}
renderPlot({
state_yr = tornados_shiny %>% 
  filter(state == input$state_selected & year==input$year_selected)
state_yr%>%
    ggplot(aes(x=month, y=injuries))+
    geom_point()+
    labs(title = "Tornados total injuries",
       x="Month", y="tornados injuries", caption = "Chuyi Sun")
})
```

# History      

## Column { .sidebar }

### Historical Data

```{r}
selectInput(inputId = "state_history", 
              label = "Select a state to see the trend of average tornodao magnitude:",
              choices = tornados_shiny%>% 
              pull(state) %>% 
              unique()%>%
              sort(), 
              selected = "AL")
```

## Column 
### Average tornados magnitude {data-width=700}
```{r, fig.width=12}
renderPlot({
state_plot = tornados_shiny %>% 
  filter(state == input$state_history)%>%
  group_by(year)%>%
  summarise(mag_mean=mean(magnitude, na.rm = T))

state_plot%>%
  ggplot(aes(x=year, y=mag_mean))+
  geom_line()+
  labs(title = paste(input$state_history,"average tornados magnitude from 2012-2020"),
      x="years", y="average tornados magnitude", caption = "Chuyi Sun")+
  theme(plot.title=element_text(size = 15), plot.subtitle = element_text(size = 9))

  })
```


Analysis    
=====================================

```{r}
FL=tornados%>%
  filter(st=="FL")%>%
  group_by(yr)%>%
  select(yr, mag)%>%
  summarise(fl_mag_mean=mean(mag, na.rm = T))

```

### Plot 1 

```{r,fig.width=10}
FL%>%
  ggplot(aes(x=yr, y=fl_mag_mean))+
  geom_line()+
  labs(title = "Florida average tornados magnitude from 1950-2020",
       subtitle = "There is a decreasing trend from 1950 to 2020. 
The lowest period is around 2000.",  x="years", y="average tornados magnitude", caption = "Chuyi Sun")+
  theme(plot.title=element_text(size = 15), plot.subtitle = element_text(size = 9))

```

### Plot 2 
```{r, fig.height=12, fig.width=10}
state_summary=tornados%>%
  mutate(tz=as.factor(tz))%>%
  mutate(st=as.factor(st))%>%
  select(-c("slat", "slon", "elat", "elon"))%>%
  group_by(st,mo)%>%
  arrange(mag)%>%
  summarise(state_mean=mean(mag, na.rm = T),.groups = "drop")
state_summary%>%
  ggplot(aes(x=st, y=state_mean))+
  geom_point()+
  facet_wrap(~mo, nrow = 6, scales = "free")+ 
  labs(title = "Average tornados magnitude of US states in each months",
       subtitle = "Except for September, other months tend to have similar 
average tornado magnitude. Alaska almost never has a tornado.", x="States", y="average tornados magnitude", caption = "Chuyi Sun")+
  geom_text(aes(label=st),size=2,vjust = -0.5)+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1, size = 5))

```

### Plot 3  

```{r,fig.width=10}
tornados%>%
  mutate(stage=case_when(yr>1949 & yr<1975 ~ "1950-1974", yr>1974 & yr<2000 ~ "1975-1999",yr>1999 & yr<2025 ~ "2000-2024"))%>%
  select(stage, inj)%>%
  group_by(stage)%>%
  summarise(stage_sum=sum(inj))%>%
  ggplot(aes(x=stage, y=stage_sum))+
  geom_bar(stat="identity")+
  labs(title = "Total injuries in tornados in every 24 years",
       subtitle = "Between 1950-1974, there were a lot of injuries, 
and 2000-2024 had the lowest injuries. There is a decreasing trend over time.", x="States", y="average tornados magnitude", caption = "Chuyi Sun")

```

### Summary  

We can see a decreasing trend in the magnitude of the tornados, therefore the trend of injuries is also decreasing. Besides, the reduction of injuries might be due to the development of prevention. Looking at the spread of the tornado in each month, except for September, other months tend to have similar average tornado magnitude.

###  Functions used 

##### Dplyr  

* mutate
* group_by
* filter
* select
* case_when
* summarise


##### ggplot2  

* geom_line
* geom_point
* geom_bar
* geom_text
* facet_wrap

##### purrr 

* map_chr
* map 


## statics content can overlap with data analysis 



